---
title: "CompStat/R - Paper 1"
author: "Group 2: Carlo Michaelis, Patrick Molligo, Lukas Ruff"
date: "11 May 2016"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Part I

### 1. What are the atomic vector types in R? Explain which value they can take and give an example!

There are six “atomic” (or “basic”) vector types in R:

* character: Text, i.e. string variables.
* numeric: Real numbers, i.e. float variables.
* integer: Integers, i.e. values in  $\mathbb{Z}$.
* complex: Complex numbers, i.e. a pair of values with a real and imaginary part.
* logical: Boolean variables, i.e. either 1 (`TRUE`) or 0 (`FALSE`).
* raw: A raw vector contains fixed-length sequences of bytes.

**Examples**
```{r ExAtomicVectorTypes}
a <- c("blue", "red", "yellow")     ## character
b <- c(pi, exp(1), 0, 1)            ## numeric
c <- 1:10                           ## integer
d <- c(0+1i, 1+1i)                  ## complex
e <- c(TRUE, FALSE)                 ## logical
f <- raw(length = 3L)               ## raw
```

It is important to note, that a vector can only contain elements of the same type. We can check the type of an object using the `class`-function.

