---
title: "CompStat/R - Paper 2"
author: "Group 2: Carlo Michaelis, Patrick Molligo, Lukas Ruff"
date: "21 June 2016"
fontsize: 11
lof: false
graphics: true
documentclass: article
output:
  pdf_document:
    latex_engine: pdflatex
    keep_tex: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      message = FALSE)
```


## Part I: Functions

### Functions I

Below we define a function `dropNa` which, given an atomic vector `x` as an argument, returns `x` after removing missing values.

```{r dropNa}
dropNa <- function(x) {
  # Expects an atomic vector as an argument and returns it without missing
  # values
  # 
  # Args:
  #   x: atomic vector
  #
  # Returns:
  #   The atomic vector x without missing values
  
  # To remove the NAs, we use simple logical subsetting
  y <- x[!is.na(x)]
  
  # Return y
  y
}
```

Let's test our implementation with the following line of code:

```{r dropNaTest}
all.equal(dropNa(c(1, 2, 3, NA, 1, 2, 3)), c(1, 2, 3, 1, 2, 3))
```

As we can see from this positive test, our implementation was successful.

### Functions II

#### Part I

Below we define a function `meanVarSdSe` which, given a numeric vector `x` as an argument, returns the mean, the variance, the standard deviation and the standard error of `x`.

```{r meanVarSdSe}
meanVarSdSe <- function(x) {
  # Expects a numeric vector as an argument and returns the mean,
  # the variance, the standard deviation and the standard error
  # 
  # Args:
  #   x: numeric vector
  #
  # Returns:
  #   a numerical vector containing mean, variance, standard deviation
  #   and standard error of x
  
  # We check if x is numeric vector
  # If not: stop and throw error
  if( !is.numeric(x) ) {
    stop("Argument needs to be numeric.")
  }
  
  # Create vector object
  y <- vector()
  
  # Calculate mean, variance, standard deviation and standard error
  # and save it in y
  y[1] <- mean(x)
  y[2] <- var(x)
  y[3] <- sd(x)
  y[4] <- y[3]/sqrt(length(x))
  
  # Set names to vector entries
  names(y) <- c("mean", "var", "sd", "se")
  
  # Return the numeric vector y
  y
}
```

To test the function, we define a numeric vector, which contains numbers from $1$ to $100$, and use it as an argument for our function `meanVarSdSe`:

```{r meanVarSdSeTest}
x <- 1:100
meanVarSdSe(x)
```

Finally we can confirm that the result is of class `numeric`:

```{r meanVarSdSeTest2}
class(meanVarSdSe(x))
```

#### Part II

Now we will have a look at the case below. We would expect that the function will return a vector with `NA`s:

```{r meanVarSdSeNAProblem}
x <- c(NA, 1:100)
meanVarSdSe(x)
```

The reason for the result is that the functions `mean()`, `var()` and `sd()` use `na.rm = FALSE` as default, which means that missing values are not removed. If the vector `x` contains a missing value, the `mean()` function (as well as `var()` and `sd()`) will just return `NA` to inform about missing values. In the case of calculating standard error we use the result from our `sd()` function and calculate an `NA` value with some other numeric values, which will ultimately result in `NA` again.

To solve the problem, we can add `na.rm = TRUE` to these three functions. To make this optional, we will improve the `meanVarSdSe` function from above as follows:

```{r meanVarSdSeImprove}
meanVarSdSe <- function(x, ...) {
  # Expects a numeric vector and flag to handle missing values as an argument
  # and returns the mean, the variance, the standard deviation
  # and the standard error
  # 
  # Args:
  #   x: numeric vector, na.rm: boolean
  #
  # Returns:
  #   a numerical vector containing mean, variance, standard deviation
  #   and standard error of x
  
  # We check if x is numeric vector
  # If not: stop and throw error
  if( !is.numeric(x) ) {
    stop("Argument needs to be numeric.")
  }
  
  # Create vector object
  y <- vector()
  
  # Calculate mean, variance, standard deviation and standard error
  # and save it in y
  y[1] <- mean(x, ...)
  y[2] <- var(x, ...)
  y[3] <- sd(x, ...)
  y[4] <- y[3]/sqrt(length(x) - sum(is.na(x)))
  
  # Set names to vector entries
  names(y) <- c("mean", "var", "sd", "se")
  
  # Return the numeric vector y
  y
}
```

We define the function with an ellipse `...`. Our function can now receive multiple arguments after the first input `x`. These arguments are used in `mean()`, `var()` and `sd()`. If we want to remove missing values in all of these functions (to get a result in the case of missing values), we can pass `na.rm = TRUE` as another argument, such as here: `meanVarSdSe(x, na.rm = TRUE)`. We just have to be aware of `length(x)` in this case. If we want to have the same result as above we have to remove the sum of `NA` values from the length of `x`. Otherwise the function will calculate a different result than in Part I, because then lentgh differs.

Let's confirm the result:

```{r meanVarSdSeImproveTest}
meanVarSdSe(c(x, NA), na.rm = TRUE)
```

#### Part III

Now we will use the function `dropNa` from Functions I to deal with missing values in `meanVarSdSe`.

```{r meanVarSdSeDrop}
meanVarSdSe <- function(x) {
  # Expects a numeric vector as an argument and returns the mean,
  # the variance, the standard deviation and the standard error
  # it also removes missing values if x contains some
  # 
  # Args:
  #   x: numeric vector
  #
  # Returns:
  #   a numerical vector containing mean, variance, standard deviation
  #   and standard error of x
  
  # We check if x is numeric vector
  # If not: stop and throw error
  if( !is.numeric(x) ) {
    stop("Argument needs to be numeric.")
  }
  
  # We check if x contains missing values
  # If so: remove missing values using dropNA
  if( sum(is.na(x)) > 0 ) {
    x <- dropNa(x)
  }
  
  # Create vector object
  y <- vector()
  
  # Calculate mean, variance, standard deviation and standard error
  # and save it in y
  y[1] <- mean(x)
  y[2] <- var(x)
  y[3] <- sd(x)
  y[4] <- y[3]/sqrt(length(x))
  
  # Set names to vector entries
  names(y) <- c("mean", "var", "sd", "se")
  
  # Return the numeric vector y
  y
}
```

We used the function from Part I and added a condition which checks if we have missing values in `x`, using `is.na`. If the sum of `NA` values is greater than $0$ (i.e.,if there is one or more missing value), we use the function `dropNA` from the first exercise to remove all missing values. The remaining code of the function can remain as above in Part I.

We can confirm the result:

```{r meanVarSdSeDropTest}
meanVarSdSe(c(x, NA))
```

### Functions III

In this section we define an infix function `%or%`. This function should behave like the logical operator `|`.

```{r infixOr}
# Define infix function %or%
`%or%` <- function(a, b) {
  # Check if vector a and b is logical
  if( !(is.logical(a) & is.logical(b)) ) {
    stop("a and/or b have to be logical vectors.")
  }
  
  # Use ifelse to calculate result and return it directly
  # If the sum of the entry of vector a and the entry of vector b
  # is greater than or equal to 1, set result to TRUE, otherwise to FALSE
  ifelse(a + b >= 1,  TRUE, FALSE)
}
```

First we check if we have logical vectors. If `a` and/or `b` are not logical, we leave the function and throw an error. Otherwise we can calculate the `or` operation using the `ifelse` function and return the result directly after calculation. Inside of the `ifelse` function, the first argument checks the condition if the sum of the values `a` and `b` are greater than or equal to $1$, where `TRUE` is equal to $1$ and `FALSE` is equal to $0$.

To confirm the function, we test an example:

```{r infixOrTest}
c(TRUE, FALSE, TRUE, FALSE) %or% c(TRUE, TRUE, FALSE, FALSE)
```

## Part II: Scoping and related topics

### Scoping I

The main concept behind this exercise is the *Search Path*, which `R` uses to locate objects when called upon. In order for `R` to carry out a command or calculation, it seeks the necessary information according to a hierarchical path of *environments*. Each environment has a *parent*, to which `R` moves if the required information is not yet found. The `R` workspace is known as the *Global Environment* and also has a parent, which is the most recently loaded package. If there are no longer any loaded packages, then the search path *ends* at the final parent environment, the base package (`package:base`) which just has the empty environment as a parent.

Below we can observe the importance of the search path with a simple example:  

```{r searchPath}
# Assign numeric values to the vectors x and y in the workspace 
# which we call the global environment
x <- 5
y <- 7

f <- function() x * y
  # With no specified argument inputs, the function f follows the search path
  # and locates values for x and y in the global environment
g <- function(x = 2, y = x) x * y
  # A new environment is created within the function g, where arguments for x and y
  # are clearly defined
```

Although both functions `f` and `g` depend on values for `x` and `y`, they return different results when called:  

```{r searchPathCalls}
# Call 1
f()
# Call 2
g()
```

Beginning with function `f`, if we follow the search path we begin in the temporary local environment within the function itself. Since there is no information regarding the values of `x` and `y`, `R` moves to the parent environment, which is the global environment in this case. In the global environment, `x` takes the value of $5$ and `y` takes the value of $7$. Thus, the function returns $5 \cdot 7=35$.  

For function `g` the search path also begins in the local environment within the function itself. However, in this case there is a defined value for `x`, as well as an expression defining a value for `y` based on `x`. The search path ends and the function returns $2 \cdot 2=4$.  

By manipulating the arguments of a function, it is also possible to alter the original search path. We see this when calling the following function:

```{r searchPathAlter}
# Call 3
g(y = x)
```

Looking back at the code for function `g`, we see the two arguments `x` and `y`. When calling `g(y = x)` however, we are omitting the first argument `x`, which then defaults to the value $2$, defined in the local environment of the function.  
When we simply call `g()`, the `y = x` argument also defaults to a local value dependent on local `x`. But by inputing the argument `y = x` manually while calling, we send the search path to the global environment where `x` takes the global value of $5$. Thus the function returns $2 \cdot 5=10$.

### Scoping II

In this exercise we once again see the importance of understanding the search path and how `R` carries out tasks according to the environment in which it is working. Especially important is the *naming* of objects and functions. As discussed in the previous section, the ultimate parent environment to use is `package:base`, which contains the commonly used and most fundamental functions in `R`. Since the global environment (workspace) is separate from `package:base`, it is possible to name new objects in the workspace using previously defined functions from the base. As long as there is no overlap *within* an enviornment, nothing will be overwritten, it will just be masked. In the following example we see again why the search path is so important when defining objects:

```{r variableOrFunction}
# Define matrix t, where the number of columns is selected as 3
# and the matrix is filled row-wise
t <- matrix(1:6, ncol = 3, byrow = TRUE)

# Print matrix t
t
```

As expected, printing `t` returns a $2\times3$ matrix filled by row using the numbers one through six. Let's see what happens if we treat `t` like a function:

```{r transposeFunction}
# Print t(t), which should transpose matrix t
t(t)
```

The result is a $3\times2$ matrix filled by column using the numbers one through six. In other words, we have printed the transpose of the original matrix `t`, which we had defined in the global environment. Since `t` is a defined matrix and not a function, `R` will ignore the `t` in the global environment while searching for function `t`. `R` follows the search path from the global environment to the earlier parents and finds function `t` in `package:base`. In the base environment, the function `t()` returns the transpose of the given matrix.

### Scoping III

In the previous exercises we observed how `R` searches through a chain of environments to locate objects and information. In this next exercise, we investigate what happens when different objects are defined identically within the *same* environment. Here we are defining objects in the global environment (workspace):

```{r globalObjects}
# Define a function t in the global environment
t <- function(...) matrix(...)

# Define a matrix T using the above t function
# with the desired input arguments
T <- t(1:6, ncol = 3, byrow = TRUE)

# Print result of T
T
```

As expected, printing `T` returns a $2 \times 3$ matrix filled by row using the numbers one through six. Now let's enter `T` into the function `t`: 

```{r printFunctionT}
# Call defined function t with argument T
t(T)
```

Since `t` is a function we have defined in our workspace (global environment), `t()` takes `T` as an argument input and returns a column vector containing the numbers one through six (note: `t` and `T` are different objects, because `R` is case sensitive). The transpose function `t()` from the base environment is now masked by our own function, saved in the global environment and now is just an alias of `matrix()` (as we defined it). Since the `matrix()` function has the default value `ncol = 1` and we were not giving any argument, it creates a matrix out of the data from `T` (values $1$ to $6$) and put it in just one column.

Let's now see what would happen if we had defined `T` as `t` instead:

```{r nameMasking}
# Define a function t in the global environment
t <- function(...) matrix(...)

# We now define t as the following matrix using the t function from above
t <- t(1:6, ncol = 3, byrow = TRUE)
# Although we used the function t to define the new matrix t
# both are defined in the global environment

# Call defined function t with argument t
t(t)
```

Since two objects (independent of their type) cannot have the same name within our global environment, the new matrix `t` overwrites the original function. In our global environment, `t` is now a defined matrix and no longer a function, as it was replaced. The search path now moves down (looks earlier on the path), until it finds `t`, defined as the transpose function (seen earlier), in the base environment. It is therefore clear why we receive the same result as in the previous exercise when printing `t(t)` here.  

This entire concept can be referred to as *name masking*. We can think of the transpose function `t` in the base environment as the *original* function. Each time a new object `t` is created in later environments, the original is *masked*, but not overwritten. So if the search path is led back to the base environment, the original function can still be located. 

### Dynamic lookup

`R` searches for objects while it runs some given code, which is called *dynamic lookup*. A *well defined* function should only process local variables. In other words a function should only depend on the given arguments. This property is called *self containment*. If we use functions that are not well defined, either we risk errors or we receive inconsistent results. For illustration, we can use the following example:

```{r dynamicLookup1}
# First we remove everything from workspace
# to avoid conflicts with the code above
rm(list = ls(all.names=TRUE))

# We define a function with two arguments
# which is well defined (it just processes local variables)
f <- function(x, y = x + 1) x + y

# We set variable x to 3 and call the function where the first argument is 2
# to see that we just use the local variable
x <- 3
f(2)  # Call 1

# We set variable x to 5 and call the function where the first argument is 2
# to see that we just use the local variable again
x <- 5
f(2)  # Call 2
```

As mentioned in the comments, we first define a well defined function `f()`. The function has two arguments, where `x` has no default and `y` has a default which depends on the local `x`: `y = x + 1`. If we set a global variable `x` (i.e. in the global environment), it will not be used in the function, unless we pass it as an argument, which is not the case here. In both calls, *call 1* and *call 2*, we pass $2$ as an argument, which leads to the same result in both cases. The global variables `x` are not used at all.

```{r dynamicLookup2}
# We define a function with one argument
# which is NOT well defined (variable x is not local)
f <- function(y = x + 1) x + y

# We set variable x to 3 and overwrite the default argument with 2
# to see that we use the global variable
x <- 3
f(2)  # Call 3

# We set variable x to 5 and overwrite the default argument with 2
# to see that we use the global variable again
x <- 5
f(2)  # Call 4

# Variable x is still set to 5 and we call the function with the default argument
# to see that we use the global variable again
f()  # Call 5
```

As mentioned in the comments, we first define a function `f()` which is *not* well defined. The function has one argument `y` which has a default that depends on the global `x`. The formula for `y` is the same as before: `y = x + 1`. In this case we *must* set a global variable `x`, otherwise we will run into errors, such as `Error in f(2) : object 'x' not found`.

In *call 3* we set the argument `y` to $2$. Inside of the function the pre-defined global variable `x` is used, which is set to $3$. The function calculates `x + y` which is $3 + 2 = 5$ in this case.

In *call 4* we set the argument `y` to $2$ again. Inside of the function the pre-defined global variable `x` is used again, which is set to $5$ in this case. The function calculates `x + y` which is $5 + 2 = 7$ in this case.

In *call 5* we use the default argument `y = x + 1`, which depends on global `x`. Inside of the function the output formula changes to `x + x + 1`, which depends on the pre-defined global variable `x` again, which is still set to $5$. The calculation $5 + 5 + 1$ results in $11$.

Although we can often get away with using functions that are not well defined, we should nevertheless always use them to avoid errors and prevent unexpected or inconsistent results.
