---
title: "CompStat/R - Paper 2"
author: "Group 2: Carlo Michaelis, Patrick Molligo, Lukas Ruff"
date: "21 June 2016"
fontsize: 11
lof: false
graphics: true
documentclass: article
output:
  pdf_document:
    latex_engine: pdflatex
    keep_tex: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      message = FALSE)
```


## Part I: Functions

### Functions I

Below we define a function `dropNa` which given an atomic vector `x` as argument, returns `x` after removing missing values.

```{r dropNa}
dropNa <- function(x) {
  # expects an atomic vector as an argument and returns it without missing
  # values
  # 
  # Args:
  #   x: atomic vector
  #
  # Returns:
  #   The atomic vector x without missing values
  
  # To remove the NAs, we use simple logical subsetting
  y <- x[!is.na(x)]
  y
}
```

Let's test our implementation with the following line of code:

```{r dropNaTest}
all.equal(dropNa(c(1, 2, 3, NA, 1, 2, 3)), c(1, 2, 3, 1, 2, 3))
```

As we can see from this positive test, our implementation was successful.

### Functions II

### Functions III


## Part II: Scoping and related topics

### Scoping I

### Scoping II

### Scoping III

### Dynamic lookup