---
title: "CompStat/R - Paper 2"
author: "Group 2: Carlo Michaelis, Patrick Molligo, Lukas Ruff"
date: "21 June 2016"
fontsize: 11
lof: false
graphics: true
documentclass: article
output:
  pdf_document:
    latex_engine: pdflatex
    keep_tex: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      message = FALSE)
```


## Part I: Functions

### Functions I

Below we define a function `dropNa` which, given an atomic vector `x` as an argument, returns `x` after removing missing values.

```{r dropNa}
dropNa <- function(x) {
  # Expects an atomic vector as an argument and returns it without missing
  # values
  # 
  # Args:
  #   x: atomic vector
  #
  # Returns:
  #   The atomic vector x without missing values
  
  # To remove the NAs, we use simple logical subsetting
  y <- x[!is.na(x)]
  
  # Return y
  y
}
```

Let's test our implementation with the following line of code:

```{r dropNaTest}
all.equal(dropNa(c(1, 2, 3, NA, 1, 2, 3)), c(1, 2, 3, 1, 2, 3))
```

As we can see from this positive test, our implementation was successful.

### Functions II

#### Part I

Below we define a function `meanVarSdSe` which, given a numeric vector `x` as an argument, returns the mean, the variance, the standard deviation and the standard error of `x`.

```{r meanVarSdSe}
meanVarSdSe <- function(x) {
  # Expects a numeric vector as an argument and returns the mean,
  # the variance, the standard deviation and the standard error
  # 
  # Args:
  #   x: numeric vector
  #
  # Returns:
  #   a numerical vector containing mean, variance, standard deviation
  #   and standard error of x
  
  # We check if x is numeric vector
  # If not: stop and throw error
  if( !is.numeric(x) ) {
    stop("Argument needs to be numeric.")
  }
  
  # Create vector object
  y <- vector()
  
  # Calculate mean, variance, standard deviation and standard error
  # and save it in y
  y[1] <- mean(x)
  y[2] <- var(x)
  y[3] <- sd(x)
  y[4] <- y[3]/sqrt(length(x))
  
  # Set names to vector entries
  names(y) <- c("mean", "var", "sd", "se")
  
  # Return the numeric vector y
  y
}
```

To test the function, we define a numeric vector, which contains numbers from $1$ to $100$, and use it as an argument for our function `meanVarSdSe`:

```{r meanVarSdSeTest}
x <- 1:100
meanVarSdSe(x)
```

Finally we can confirm that the result is of class `numeric`:

```{r meanVarSdSeTest2}
class(meanVarSdSe(x))
```

#### Part II

Now we will have a look at the case below. We would expect that the function will return a vector with `NA`s:

```{r meanVarSdSeNAProblem}
x <- c(NA, 1:100)
meanVarSdSe(x)
```

The reason for the result is that the functions `mean()`, `var()` and `sd()` use `na.rm = FALSE` as default, which means that missing values are not removed. If the vector `x` contains a missing value, the `mean()` function (as well as `var()` and `sd()`) will just return `NA` to inform about missing values. In the case of calculating standard error we use the result from our `sd()` function and calculate an `NA` value with some other numeric values, which will ultimately result in `NA` again.

To solve the problem, we can add `na.rm = TRUE` to these three functions. To make this optional, we will improve the `meanVarSdSe` function from above as follows:

```{r meanVarSdSeImprove}
meanVarSdSe <- function(x, ...) {
  # Expects a numeric vector and flag to handle missing values as an argument
  # and returns the mean, the variance, the standard deviation
  # and the standard error
  # 
  # Args:
  #   x: numeric vector, na.rm: boolean
  #
  # Returns:
  #   a numerical vector containing mean, variance, standard deviation
  #   and standard error of x
  
  # We check if x is numeric vector
  # If not: stop and throw error
  if( !is.numeric(x) ) {
    stop("Argument needs to be numeric.")
  }
  
  # Create vector object
  y <- vector()
  
  # Calculate mean, variance, standard deviation and standard error
  # and save it in y
  y[1] <- mean(x, ...)
  y[2] <- var(x, ...)
  y[3] <- sd(x, ...)
  y[4] <- y[3]/sqrt(length(x) - sum(is.na(x)))
  
  # Set names to vector entries
  names(y) <- c("mean", "var", "sd", "se")
  
  # Return the numeric vector y
  y
}
```

We define the function with an ellipse `...`. Our function can now receive multiple arguments after the first input `x`. These arguments are used in `mean()`, `var()` and `sd()`. If we want to remove missing values in all of these functions (to get a result in the case of missing values), we can pass `na.rm = TRUE` as another argument, such as here: `meanVarSdSe(x, na.rm = TRUE)`. We just have to be aware of `length(x)` in this case. If we want to have the same result as above we have to remove the sum of `NA` values from the length of `x`. Otherwise the function will calculate a different result than in Part I, because then lentgh differs.

Let's confirm the result:

```{r meanVarSdSeImproveTest}
meanVarSdSe(c(x, NA), na.rm = TRUE)
```

#### Part III

Now we will use the function `dropNa` from Functions I to deal with missing values in `meanVarSdSe`.

```{r meanVarSdSeDrop}
meanVarSdSe <- function(x) {
  # Expects a numeric vector as an argument and returns the mean,
  # the variance, the standard deviation and the standard error
  # it also removes missing values if x contains some
  # 
  # Args:
  #   x: numeric vector
  #
  # Returns:
  #   a numerical vector containing mean, variance, standard deviation
  #   and standard error of x
  
  # We check if x is numeric vector
  # If not: stop and throw error
  if( !is.numeric(x) ) {
    stop("Argument needs to be numeric.")
  }
  
  # We check if x contains missing values
  # If so: remove missing values using dropNA
  if( sum(is.na(x)) > 0 ) {
    x <- dropNa(x)
  }
  
  # Create vector object
  y <- vector()
  
  # Calculate mean, variance, standard deviation and standard error
  # and save it in y
  y[1] <- mean(x)
  y[2] <- var(x)
  y[3] <- sd(x)
  y[4] <- y[3]/sqrt(length(x))
  
  # Set names to vector entries
  names(y) <- c("mean", "var", "sd", "se")
  
  # Return the numeric vector y
  y
}
```

We used the function from Part I and added a condition which checks if we have missing values in `x`, using `is.na`. If the sum of `NA` values is greater than $0$ (i.e.,if there is one or more missing value), we use the function `dropNA` from the first exercise to remove all missing values. The remaining code of the function can remain as above in Part I.

We can confirm the result:

```{r meanVarSdSeDropTest}
meanVarSdSe(c(x, NA))
```

### Functions III

In this section we define an infix function `%or%`. This function should behave like the logical operator `|`.

```{r infixOr}
# Define infix function %or%
`%or%` <- function(a, b) {
  # Check if vector a and b is logical
  if( !(is.logical(a) & is.logical(b)) ) {
    stop("a and/or b have to be logical vectors.")
  }
  
  # Use ifelse to calculate result and return it directly
  # If the sum of the entry of vector a and the entry of vector b
  # is greater than or equal to 1, set result to TRUE, otherwise to FALSE
  ifelse(a + b >= 1,  TRUE, FALSE)
}
```

First we check if we have logical vectors. If `a` and/or `b` are not logical, we leave the function and throw an error. Otherwise we can calculate the `or` operation using the `ifelse` function and return the result directly after calculation. Inside of the `ifelse` function, the first argument checks the condition if the sum of the values `a` and `b` are greater than or equal to $1$, where `TRUE` is equal to $1$ and `FALSE` is equal to $0$.

To confirm the function, we test an example:

```{r infixOrTest}
c(TRUE, FALSE, TRUE, FALSE) %or% c(TRUE, TRUE, FALSE, FALSE)
```

## Part II: Scoping and related topics

### Scoping I

The main concept behind this exercise is the *Search Path*, which `R` uses to locate objects when called upon. In order for `R` to carry out a command or calculation, it seeks the necessary information according to a hierarchical path of "environments". Each environment has a "parent" environment, to which `R` moves if the required information is not yet found. The `R` workspace is known as the *Global Environment* and also has a parent, which is either the most recently loaded package or the base package (`package:base`), also known as the "end" of the search path.  

Below we can observe the importance of the search path with a simple example:  

```{r searchPath}
# We assign numeric values to the vectors x and y in the workspace 
# which we call the global environment
x <- 5
y <- 7

f <- function() x * y
  # With no specified argument inputs, the function f follows the search path
  # and locates values for x and y in the global environment
g <- function(x = 2, y = x) x * y
  # A new environment is created within the function g, where arguments for x and y
  # are clearly defined
```

Although both functions `f` and `g` depend on values for `x` and `y`, they return different results when called:  

```{r searchPathCalls}
f()
g()
```

Beginning with function `f`, if we follow the search path we begin in the environment within the function itself. Since there is no information regarding the values of `x` and `y`, `R` moves to the parent environment, which is the global environment in this case. In the global environment, `x` takes the value of $5$ and `y` takes the value of $7$. Thus, the function returns $5*7=35$.  
For function `g` the search path also begins in the environment within the function itself. However, in this case there is a defined value for `x`, as well as an expression defining a value for `y` based on `x`. The search path ends and the function returns $2*2=4$.  

By manipulating the arguments of a function, it is also possible to alter the original search path. We see this when calling the following function:

```{r searchPathAlter}
g(y = x)
```

Looking back at the code for function `g`, we see the two arguments `x` and `y`. When calling `g(y = x)` however, we are ignoring the first argument, which then defaults to the value $2$. When we simply call `g()`, the `y = x` argument also defaults to a value dependent on `x`. But by inputing `y = x` manually, we send the search path to the global environment where `x` takes the value of $5$. Thus the function returns $2*5=10$.   



### Scoping II

### Scoping III

### Dynamic lookup