---
title: "CompStat/R - Paper 2"
author: "Group 2: Carlo Michaelis, Patrick Molligo, Lukas Ruff"
date: "21 June 2016"
fontsize: 11
lof: false
graphics: true
documentclass: article
output:
  pdf_document:
    latex_engine: pdflatex
    keep_tex: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      message = FALSE)
```


## Part I: Functions

### Functions I

Below we define a function `dropNa` which given an atomic vector `x` as argument, returns `x` after removing missing values.

```{r dropNa}
dropNa <- function(x) {
  # expects an atomic vector as an argument and returns it without missing
  # values
  # 
  # Args:
  #   x: atomic vector
  #
  # Returns:
  #   The atomic vector x without missing values
  
  # To remove the NAs, we use simple logical subsetting
  y <- x[!is.na(x)]
  
  # Return y
  y
}
```

Let's test our implementation with the following line of code:

```{r dropNaTest}
all.equal(dropNa(c(1, 2, 3, NA, 1, 2, 3)), c(1, 2, 3, 1, 2, 3))
```

As we can see from this positive test, our implementation was successful.

### Functions II

#### Part I

Below we define a function `meanVarSdSe` which given a numeric vector `x` as argument, returns the mean, the variance, the standard deviation and the standard error of `x`.

```{r meanVarSdSe}
meanVarSdSe <- function(x) {
  # expects a numeric vector as an argument and returns the mean,
  # the variance, the standard deviation and the standard error
  # 
  # Args:
  #   x: numeric vector
  #
  # Returns:
  #   a numerical vector containing mean, variance, standard deviation
  #   and standard error of x
  
  # First we check if we have a numerical vector
  # If not: stop and throw error
  if( !is.numeric(x) )
    stop("Argument need to be numeric.")
  
  # Create vector object
  y <- vector()
  
  # Calculate mean, variance, standard deviation and standard error
  # and save it in y
  y[1] <- mean(x)
  y[2] <- var(x)
  y[3] <- sd(x)
  y[4] <- y[3]/sqrt(length(x))
  
  # Set names to vector entries
  names(y) <- c("mean", "var", "sd", "se")
  
  # Return the numeric vector y
  y
}
```

To test the function, we define a numeric vector, which contains numbers from $1$ to $100$ and use it as an argument for our function `meanVarSdSe`:

```{r meanVarSdSeTest}
x <- 1:100
meanVarSdSe(x)
```

Finally we can confirm, that the result is of type `numeric`:

```{r meanVarSdSeTest2}
class(meanVarSdSe(x))
```

#### Part II

#### Part III

### Functions III


## Part II: Scoping and related topics

### Scoping I

### Scoping II

### Scoping III

### Dynamic lookup